Compositional Certified Resource Bounds

### 2014-06-30 paper/main.tex
title?
updates to intro and abstract

% \title{Automatic Amortized Analysis for Imperative Code}
\title{Compositional Quantitative Resource Analysis of C Programs}
% Compositional Resource-Bound Certification for C Programs

@@

  This paper describes a novel quantitative resource analysis
  framework that tackles these challenges for C programs.  The
  analysis framework consists of three parts.  First, the runtime
  resource cost of programs is formalized with a parametric cost
  semantics for CompCert Clight.  Second, a quantitative Hoare logic
  that can be used for interactive bound development is implement and
  proved sound with respect to the cost semantics in the Coq Proof
  Assistant.  Third, an automatic amortized resource analysis for C
  programs that computes derivations in the quantitative Hoare logic
  is developed and implemented.  It is the first automatic amortized
  analysis that can derive bounds that depend on negative integers and
  differences between numbers.  Both, the quantitative logic and the
  automatic amortized analysis are naturally compositional and can be
  combined to semi-automatically derive global resource bounds.

  The practicability of the analysis framework is demonstrated with an
  experimental evaluation.  The logic is used to derive XXX A comparison
  with other automatic tools shows that the constant factors in the
  bounds that are derived by the automatic amortized analysis are
  often more precise. XXX

\end{abstract}

@@

  analysis is based on type-based amortized resource analysis for
  functional programs~\cite{Jost03,HoffmannAH12} and computes

@@

  solved by an off-the-shelf LP solver.

@@

  recursion patterns and that the derived constant factors in the

@@

By compiling these bounds together with the C program, we derive

@@

Our \emph{first contribution} is a generalization of the quantitative
Hoare logic for CompCert Clight that enables us to derive resource
bounds that are parametric in the resource of interest.  The
quantitative logic is implemented in Coq and proved sound with respect
to a small-step operational cost semantics for Clight that counts a
constant resource consumption for each atomic execution step.  The
constant resource consumption at each step is parametric in
user-definable cost metric and can be negative to express the release
of resources (as needed, e.g., to reason about memory usage).
Furthermore, a specific \emph{tick} function can be inserted in the
source code to further customize the cost model.  In the
implementation of the logic, we use a shallow embedding in Coq which
makes the logic very flexible.  There is practically no limitation on
the format of the derived resource bounds since they can be any
function that is definable in Coq.  The derived bounds can also be
parametric in a set of cost metrics or specifically apply to one fixed
cost metric.

The precision and flexibility of the quantitative Hoare logic provide
a ample foundation for reasoning about resource consumption of Clight
programs.  The logic is naturally compositional and the quantitative
Hoare triples can be used to specify resource cost of library
functions or to describe resource contracts between different
functions.  However, reasoning about quantitative properties can often
be more tedious then reasoning about intentional properties.
Consequently, automation is inevitable to derive resource bounds for
large code bases like CertiKOS.  Such an automation was easily
achievable in the case of constant stack bounds for a code base that
does not contain recursive functions.  In general however, resource
bounds have to be parametric in the arguments of a function and depend
on the number of loop iterations and (recursive) function calls that
are performed by the function.

For functional programs, there exist resource analysis systems that
are based on type-based amortized resource analysis~\cite{Jost03,
  HoffmannAH12} and can automatically derive complex polynomial bounds
with tight constant factors.  Type-based amortized resource analysis
has been our main inspiration in the design of the quantitative Hoare
logic and is therefore a natural candidate to automate the reasoning.
However, it is a well-known problem that type-based amortized resource
analysis works well for functional programs with pattern matching but
not for C-like programs with control flow that depends integer
arithmetic.  For example, it is a long-time open problem how to use
type-based amortized analysis to derive bounds for functions whose
resource consumption depends on (possibly negative) integers and
non-sequential control flow as introduced by \code{break} and
\code{return} statements.  Unfortunately, these are the functions with
non-constant resource consumption that are most common in CertiKOS and
other system software.

Our \emph{second contribution} is an automatic amortized resource
analysis for CompCert Clight that computes derivations in the
quantitative Hoare logic.  It is the first automatic amortized
resource analysis that can derive bounds that depend on negative
integers and handles programs with mutually recursive as well as
\code{break} and \code{return} statements.  It targets programs that
are common in system software, that is, programs whose resource
consumption can be described as a function of sizes $|[x,y]| =
\max(0,y-x)$ of intervals of integer variables.  Such bounds arise for
instance from standard \emph{for loops} \code{for (i = x; i+K \leq y;
  i=i+K)} where the step-size $K>0$ is a constant.

The use of interval sizes in potential functions (an therefore in
resource bounds) is the main innovation that makes the analysis
practical for C.
%
Despite the apparent simplicity of the new analysis system, it is able
to reproduce results from the literature (e.g,
SPEED~\cite{GulwaniMC09}) that have been obtained using sophisticated
abstract--interpretation base methods.  In contrast with abstract
interpretation--based methods, our technique does not require any
fixpoint computations to obtain loop invariants.  The mechanism we
designed is able to leverage local assertions such as $x < y$ that we
collect along the branching points of the program to obtain global
resource invariants.  We achieve this by generating a linear
constraint system that reflects resource cost and size changes of
integer variables in the program.  A solution of the linear program is
a resource usage bound for the C program.  % We can naturally, handle
% advanced control flow such as break, return, iteration over negative
% integer variables, and mutually-recursive functions.

Following the development steps of automatic amortized analysis for
functional programs~\cite{Jost03,HoffmannH10}, we deliberately
restrict our self to linear bounds in the automatic analysis (there a
no restrictions for the manually-derived bounds in the logic).  More
specifically, bounds have the form $\sum_{a,b} q_{(a,b)} |[a,b]|$
where $a$ and $b$ are integer variables or constants.  The reason is
mainly clarity of presentation, since an extension to multivariate
resource polynomials~\cite{HoffmannAH11} would make the inference rules considerably
more involved and should better be described separately.  However, we
developed the linear inference system so that the extension to
polynomial bounds shall work smoothly.

To the best of our knowledge, this article presents the first resource
analysis framework for C that makes it possible to combine non-trivial
automatically derived bounds with interactively derived bounds in a
prove system that produces verifiable certificates for the bounds.
Our approach complements existing work since it provides a semantic
foundation for the computation of bounds while still providing support
for automation that sometimes goes beyond the capabilities of existing
techniques.  Both the automatic amortized analysis and the
quantitative logic are naturally compositional and describe the
resource behavior of code fragments without referring to the source
code.  As a result, we naturally derive global bounds that are
functions of the input parameters of the program.  Another unique
feature of our system is, that it can naturally handle resources like
memory that may become available during the execution.

We have evaluated the quantitative logic and the automatic analysis
with system code and examples from the literature. \pref{app:cat}
contains more than $30$ (XXX) challenging loop and recursion patterns
that we collected from open source software and the literature.  Our
analysis can find asymptotically tight bounds for $30$ all but 3 of
these patterns and in most cases the derived constant factors are
tight.  We compared our automatic analyzer to other resource-analysis
systems.  We selected $5$ examples (see Section XXX) for which we
derived linear bounds and translated them to term-rewriting systems by
hand in order to run them with the tools KoAT, PUBS, and Rank.  The
tools were not able to find linear bounds for the examples.  LOOPUS is
able to derive a bound on XXX of the examples.  But in contrast to our
analysis, the constant factors are not tight.  We also contacted the
authors of SPEED but were not able to obtain a tool to perform a
comparison.

To show the practicability of our approach we developed a case study
with different real-time schedulers implemented in C.  To perform a
schedulability analysis that takes into account the overhead of the
scheduler we derived time bounds for context switching and
initialization of the scheduler.  We then computed time bounds for a
set of given user task and took account the previously computed
scheduler overhead to determine whether the scheduling requirements of
all user tasks can be met.

With the intention, of making the material easily accessible we first
informally describe the new automatic amortized analysis
(\pref{sec:inform}).  We then formalize the intuition with the
operational cost semantics an the inference rules for the amortized
analysis (\pref{sec:aa}).  In \pref{sec:logic} we describe the
quantitative Hoare logic and its soundness proof that is formalized in
Coq.  In \pref{sec:inter} we prove the soundness of the automatic
amortized analysis by showing that derivation using the inference
rules can be seen as derivations in the quantitative logic.  We also
show how to integrate hand-derived bounds with automatically-derived
bounds.  \pref{sec:exper} contains the results of the experimental
evaluation and \pref{sec:related} describes related research.

>>>>

\title{Compositional Certified Resource Bounds}
% Certified Resource Bounds for C
% Compositional Certified Resource Bounds for C
% Compositional Quantitative Resource Analysis of C Programs
% Compositional Resource-Bound Certification for C Programs

@@

  This paper describes a novel quantitative resource analysis
  framework that tackles these challenges for C programs.  The
  analysis framework consists of three parts.  First, the runtime
  resource consumption of programs is formalized with a parametric
  cost semantics for CompCert Clight.  Second, a quantitative Hoare
  logic that can be used for interactive bound development is
  implement and proved sound with respect to the cost semantics in the
  Coq Proof Assistant.  The logic is very expressive and a shallow
  embedding enables a derived bound to be any function that is
  definable in Coq.  Third, an automatic amortized resource analysis
  for C programs that computes derivations in the quantitative Hoare
  logic is developed and implemented.  It is the first automatic
  amortized analysis that can derive bounds that depend on negative
  integers and differences between numbers, which is crucial to handle
  typical systems code.  Both, the quantitative logic and the
  automatic amortized analysis are naturally compositional and can be
  combined to semi-automatically derive global resource bounds.

  The practicability of the analysis framework is demonstrated with an
  experimental evaluation.  The expressivity of the logic is
  demonstrated by manually deriving customized bounds that are
  tailored to specific algorithms.  A comparison of the automatic
  amortized analysis with other automatic tools on 30 challenging
  examples from the literature and open-source software shows that the
  derived constant factors in our bounds are often more precise.
\end{abstract}

@@

  analysis is inspired by type-based amortized resource analysis for
  functional programs~\cite{Jost03,HoffmannAH12}.  It computes

@@

  solved by an off-the-shelf LP solver and does not require any
  fixpoint computations to obtain loop invariants.

@@

  recursion patterns, and that the derived constant factors in the

@@

By compiling these bounds together with the C program, we have derived

@@

Our \emph{first contribution} is a generalization of the quantitative
Hoare logic for CompCert Clight that enables us to derive resource
bounds that are parametric in the resource of interest.  The
quantitative logic is implemented in Coq and proved sound with respect
to a small-step operational cost semantics for Clight that counts a
constant resource consumption for each atomic execution step.  The
resource consumption at each step is parametric in a user-definable cost
metric and can be negative to express the release of resources (as
needed, e.g., to reason about memory usage).  Furthermore, a specific
\emph{tick} function can be inserted in the source code to further
customize the cost model.  In the implementation of the logic, we use
a shallow embedding in Coq which makes the logic very flexible.  There
is practically no limitation on the format of the derived resource
bounds since they can be any function that is definable in Coq.  The
derived bounds can also be parametric in a set of cost metrics or
specifically apply to one fixed cost metric.

The precision and expressivity of the quantitative Hoare logic provide
an ample foundation for reasoning about resource consumption of Clight
programs.  The logic is naturally compositional and the derived
\emph{quantitative Hoare triples} can be used to specify the resource
consumption of library functions or to describe resource contracts
between different functions.  However, reasoning about quantitative
properties can often be more tedious then reasoning about intentional
properties.  Consequently, automation is inevitable to derive resource
bounds for large code bases like CertiKOS.  Such an automation was
easily achievable in the case of constant stack bounds for a code base
that does not contain recursive functions~\cite{veristack14}.  In
general however, resource bounds have to be parametric in the
arguments of a function and depend on the number of loop iterations
and (recursive) function calls that are performed by the function.

For functional programs, there exist resource analysis systems---based
on type-based amortized resource analysis~\cite{Jost03,
  HoffmannAH12}---that can automatically derive complex polynomial
bounds with tight constant factors.  This type-based amortized
resource analysis has been our main inspiration in the design of the
quantitative Hoare logic and is therefore a natural candidate to
automate the reasoning.  However, it is a well-known problem that
type-based amortized resource analysis works well for functional
programs with pattern matching but not for C-like programs with
control flow that depends integer arithmetic.  For example, it is a
long-time open problem how to use type-based amortized analysis to
derive bounds for functions whose resource consumption depends on
(possibly negative) integers and non-sequential control flow as
introduced by \code{break} and \code{return} statements.
Unfortunately, these are the functions with non-constant resource
consumption that are most common in CertiKOS and other system
software.

Our \emph{second contribution} is an automatic amortized resource
analysis for CompCert Clight that computes derivations in the
quantitative Hoare logic.  It is the first automatic amortized
resource analysis that can derive bounds that depend on negative
integers and handles programs with mutually recursive as well as
\code{break} and \code{return} statements.  It targets programs that
are common in system software, that is, programs whose resource
consumption can be described as a function of sizes $|[x,y]| =
\max(0,y-x)$ of intervals of integer variables.  Such bounds arise for
instance from standard \emph{for loops} \code{for (i = x; i+K \leq y;
  i=i+K)} where the step-size $K>0$ is a constant.
%
The use of interval sizes in potential functions (an therefore in
resource bounds) is the main innovation that makes the analysis
practical for C.

Despite the apparent simplicity of the new analysis system, it is able
to reproduce results from the literature (e.g,
SPEED~\cite{GulwaniMC09}) that have been obtained using sophisticated
abstract--interpretation base methods.  In contrast with abstract
interpretation--based methods, our technique does not require any
fixpoint computations to obtain loop invariants.  The mechanism we
designed is able to leverage local assertions such as $x < y$ that we
collect along the branching points of the program to obtain global
resource invariants.  We achieve this by generating a linear
constraint system that reflects resource cost and size changes of
integer variables in the program.  A solution of the linear program
immediately yields a resource usage bound for the C
program.  % We can naturally, handle
% advanced control flow such as break, return, iteration over negative
% integer variables, and mutually-recursive functions.

Following the development steps of automatic amortized analysis for
functional programs~\cite{Jost03,HoffmannH10}, we deliberately
restrict our self to linear bounds in the automatic analysis (there
are no restrictions for the manually-derived bounds in the logic).
More specifically, bounds have the form $\sum_{a,b} q_{(a,b)} |[a,b]|$
where $a$ and $b$ are integer variables or constants.  The reason for
our focus on linear bounds is mainly clarity of presentation.  We
already experimented with an extension to multivariate resource
polynomials~\cite{HoffmannAH11} but this would make the inference
rules considerably more involved and should better be described
separately.  However, we developed the linear inference system so that
the extension to polynomial bounds shall work smoothly.

To the best of our knowledge, this article presents the first resource
analysis framework for C that makes it possible to combine non-trivial
automatically derived bounds with interactively derived bounds in a
proof system that produces verifiable certificates for the bounds.
Our approach complements existing work since it provides a semantic
foundation for the computation of bounds while still providing support
for automation that sometimes goes beyond the capabilities of existing
techniques.  Both the automatic amortized analysis and the
quantitative logic are naturally compositional and describe the
resource behavior of code fragments without referring to the source
code.  As a result, we naturally derive global bounds that are
functions of the input parameters of the program.  Another unique
feature of our system is, that it can naturally handle resources like
memory that may become available during the execution.

We have evaluated the automatic analysis with system code and examples
from the literature.  \pref{app:cat} contains more than $30$
challenging loop and recursion patterns that we collected from open
source software and the literature.  Our analysis can find
asymptotically tight bounds for $30$ all but 3 of these patterns and
in most cases the derived constant factors are tight.  To compare our
automatic analyzer with existing techniques, we tested our examples
with the tools KoAT~\cite{BrockschmidtEFFG14},
PUBS~\cite{AlbertAGPZ12}, Rank~\cite{AliasDFG10}, and
LOOPUS~\cite{SinnZV14}.  We also contacted the authors of
SPEED~\cite{GulwaniMC09} but were able to obtain a tool to perform a
comparison.  Our experiments shows that the bounds that we derive are
often more precise than those derived by existing tools.  Only
LOOPUS~\cite{SinnZV14}, which also uses amortization techniques, is
able to achieve a similar precision.

Several micro benchmarks demonstrate the practicability and
expressiveness of the quantitative Hoare logic.  For example, we
derive a logarithmic bound for a binary search function, a bound that
depends on the contents of an array to describe the exact cost of a
function that finds the maximal element in an array, and a linear
bound that amortizes the cost of $n$ successive increments to a binary
counter.  XXX In a larger case study, we show how the quantitative
logic can be used to derive the amortized cost for data-structure
operations for a table that expands or shrinks depending on the load
factor.

% To show the practicability of our approach we developed a case study
% with different real-time schedulers implemented in C.  To perform a
% schedulability analysis that takes into account the overhead of the
% scheduler we derived time bounds for context switching and
% initialization of the scheduler.  We then computed time bounds for a
% set of given user task and took account the previously computed
% scheduler overhead to determine whether the scheduling requirements of
% all user tasks can be met.

% With the intention, of making the material easily accessible we first
% informally describe the new automatic amortized analysis
% (\pref{sec:inform}).  We then formalize the intuition with the
% operational cost semantics an the inference rules for the amortized
% analysis (\pref{sec:aa}).  In \pref{sec:logic} we describe the
% quantitative Hoare logic and its soundness proof that is formalized in
% Coq.  In \pref{sec:inter} we prove the soundness of the automatic
% amortized analysis by showing that derivation using the inference
% rules can be seen as derivations in the quantitative logic.  We also
% show how to integrate hand-derived bounds with automatically-derived
% bounds.  \pref{sec:exper} contains the results of the experimental
% evaluation and \pref{sec:related} describes related research.
